Convergence rate of MAP estimates \\ for the exponential family

### 2021-10-11 aistats.tex
Clarify that MLE is not optim

\section{Open Problem}
Given a distribution $\cD$, our objective is the expected log-likelihood 
\begin{align}
	f(\nat) = \expect[X \sim \cD]{-\log p(X \cond \nat)} 
	 = \logpart(\nat) - \langle \E[T] , \nat \rangle
\end{align}
Up to a constant, it is also the KL  $f(\nat) = \KL(\cD || \nat) := \KL(\cD || p(\cdot \cond \nat))$.
Note that $\cD$ does not have to be part of the exponential family.
Let $\nat^* := \argmin f(\nat)$.
Assuming $\nat^*$ exists within $\Theta$ means $\cD$ is such that $\E[T] = \meanp^* \in \cM$.
Then the suboptimality on the population log-likelihood can be written with Bregman divergences
\footnote{
Since for exponential families, $\bregman(\nat ; \nat^*) = KL(\nat^* || \nat)$, this proves on the way the Pythagorean theorem of information geometry $KL(\cD || \nat) = \KL(\cD || \nat^*) + \KL(\nat^* || \nat)$.
}
\begin{align}
	 f(\nat) - f(\nat^*)
	 %&= \logpart(\nat) - \logpart(\nat^*) - \langle \E[T] , \nat - \nat^* \rangle \\
	 = \bregman(\nat ; \nat^*) 
	 = \bregmanconj(\m^* ; \m) \; .
\end{align}
Our question is: how does this suboptimality behave when $\nat$ is the MLE or the MAP ? Can we get upper-bounds in expectations ? e.g. on these quantities
\begin{align}
	\label{eq:bregmanMLE}
	\expect[\mD]{\bregmanconj \left (\E [T] ;  \inv{n}  \smallsum_i T_i \right )} \\
	\label{eq:bregmanMAP}
	\expect[\mD]{\bregmanconj \left (\E [T] ; \frac{n_0 \m_0 + \smallsum_i T_i}{n_0+n} \right )}
\end{align}
where the outer expectation is on the dataset $\mD = \{X_1, \dots, X_n \}$ sampled i.i.d from $\cD$.
More explicitly, can we write an upper bound that does not involve this expectation over the dataset.

@@

\subsection{MAP as Stochastic Mirror Descent}
\label{ssec:MAP=SMD}
The posterior~\eqref{eq:joint_likelihood} can always be written iteratively $p(\nat | \mD_n) \propto p(X_n | \nat) p(\nat | \mD_{n-1})$, so that the MAP verifies
\begin{align*}
	\hat \nat_{n}
	= \argmin_\nat - \log p(X_{n} | \nat) + (n_0 + n -1) \bregman(\nat; \hat \nat_{n-1})
    %\label{eq:SBPP}
\end{align*}
which is the update of a Stochastic Bregman Proximal Point update with step-size $\inv{n_0 + n - 1}$. Expanding $p(X_{n} | \nat)$ and  playing with the Bregman we get
\begin{align}
		\hat \nat_{n}
	= \argmin_\nat g_{n}^\top \nat + (n_0 + n) \bregman(\nat; \hat \nat_{n-1})
    \label{eq:SMD}
\end{align}
where $g_{n} := \nabla\logpart(\hat \nat_{n-1}) - T(X_n) = \hat \m_{n-1} - T_n$ is the gradient of $-\log p(X_{n} | \nat)$ evaluated at $\hat \nat_{n-1}$.
We recognize the update of Stochastic Mirror Descent (a.k.a. Stochastic Bregman Gradient) with step-size $\lr_n = \inv{n_0 + n}$ -- a stochastic gradient step in the dual $\hat \m_n = \hat \m_{n-1} - \lr_n g_n$.
This parallel means it would be possible to obtain a finite convergence rate for the  MAP from an analysis of these algorithms.
We give a special focus on this approach in Section~\ref{sec:optimization}.

>>>>

\section{Open Problem}
Given a distribution $\cD$, our objective is the expected log-likelihood 
\begin{align}
	f(\nat) = \expect[X \sim \cD]{-\log p(X \cond \nat)} 
	 = \logpart(\nat) - \langle \E[T] , \nat \rangle
\end{align}
Note that $\cD$ does not have to be part of the exponential family.
Let $\nat^* := \argmin f(\nat)$.
Assuming $\nat^*$ exists within $\Theta$ means $\cD$ is such that $\E[T] = \meanp^* \in \cM$.
Then the suboptimality on the population log-likelihood can be written with Bregman divergences
\footnote{
Since $f$ is also the KL (up to a constant)  $f(\nat) + \cst = \KL(\cD || \nat) := \KL(\cD || p(\cdot \cond \nat))$, 
and for exponential families $\bregman(\nat ; \nat^*) = KL(\nat^* || \nat)$, we recover the Pythagorean theorem of information geometry $KL(\cD || \nat) = \KL(\cD || \nat^*) + \KL(\nat^* || \nat)$.
}
\begin{align}
	 f(\nat) - f(\nat^*)
	 %&= \logpart(\nat) - \logpart(\nat^*) - \langle \E[T] , \nat - \nat^* \rangle \\
	 = \bregman(\nat ; \nat^*) 
	 = \bregmanconj(\m^* ; \m) \; .
\end{align}
Our question is: how does this suboptimality behave when $\nat$ is the MLE or the MAP ? Can we get upper-bounds in expectations ? e.g. on these quantities
\begin{align}
	\label{eq:bregmanMLE}
	\expect[\mD]{\bregmanconj \left (\E [T] ;  \inv{n}  \smallsum_i T_i \right )} \\
	\label{eq:bregmanMAP}
	\expect[\mD]{\bregmanconj \left (\E [T] ; \frac{n_0 \m_0 + \smallsum_i T_i}{n_0+n} \right )}
\end{align}
where the outer expectation is on the dataset $\mD = \{X_1, \dots, X_n \}$ sampled i.i.d from $\cD$.
More explicitly, can we write an upper bound that does not involve this expectation over the dataset.

@@

\subsection{MAP as Stochastic Mirror Descent}
\label{ssec:MAP=SMD}
By Bayes rule, the posterior can always be written iteratively $p(\nat | \mD_n) \propto p(X_n | \nat) p(\nat | \mD_{n-1})$.
Plugging in~\eqref{eq:joint_likelihood}, we see that the MAP verifies
\begin{align*}
	\hat \nat_{n}
	= \argmin_\nat - \log p(X_{n} | \nat) + (n_0 + n -1) \bregman(\nat; \hat \nat_{n-1})
    %\label{eq:SBPP}
\end{align*}
which is the update of a Stochastic Bregman Proximal Point update with step-size $\inv{n_0 + n - 1}$. Expanding $p(X_{n} | \nat)$ and  playing with the Bregman we get
\begin{align}
		\hat \nat_{n}
	= \argmin_\nat g_{n}^\top \nat + (n_0 + n) \bregman(\nat; \hat \nat_{n-1})
    \label{eq:SMD}
\end{align}
where $g_{n} := \nabla\logpart(\hat \nat_{n-1}) - T(X_n) = \hat \m_{n-1} - T_n$ is the gradient of $-\log p(X_{n} | \nat)$ evaluated at $\hat \nat_{n-1}$.
We recognize the update of Stochastic Mirror Descent (a.k.a. Stochastic Bregman Gradient) with step-size $\lr_n = \inv{n_0 + n}$ -- a stochastic gradient step in the dual $\hat \m_n = \hat \m_{n-1} - \lr_n g_n$.
This parallel means we could obtain a finite convergence rate for the  MAP from an analysis of these algorithms.
In particular, it could potentially apply to a wide range of $\cD$.
Note that this optimization perspective does not hold for the MLE %($n_0=0$)
as the initial prior uniform prior may correspond to an initialization outside of $\Theta$.
We focus more on this approach in Section~\ref{sec:optimization}.
